Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2019-09-25 paper/paper.Rmd
Paper update with new analysis

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_overview.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
source('civicmine/fig_evaluationresults.R')
invisible(dev.off())

@@

**Results**: We extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed in order to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying biology of the disease by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ESR1*/*PGR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are a primary approach for testing samples for diagnostic markers (e.g. CD15 and CD30 for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increasing speed of geneome sequencing has also allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at BC Cancer identifies somatic events in the genome such as point mutations, copy number variations and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@laskin2015lessons].

@@

A number of groups have built in-house knowledgebases, which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing]. To encourage a collaborative effort, the CIViC knowledgebase (https://civicdb.org) was launched to provide a wiki-like, editable online resource where community contributed edits and additions are moderated by experts in order to maintain high quality variant curation [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer described in the peer-reviewed literature. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels and other ‘omic events. It supports four types of evidence associating biomarkers with different classes of clinical relevance (also known as evidence types).

@@

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an information extraction-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

Several groups have created knowledgebases to aid clinical interpretation of cancer genomes, many of whom have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/) . VICC is an initiative that aims to coordinate variant interpretation efforts and, to this end, has created a federated search mechanism to allow easier analysis across multiple knowledgebases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

Most of these projects focus on clinically-relevant genomic events, particularly point mutations, and provide associated clinical information tiered by different levels of evidence. Only CIViC includes RNA expression-based biomarkers. These may be of particular value for childhood cancers which are known to be ‘genomically quiet’, having accrued very few somatic mutations. Consequently, their clinical interpretation may rely more heavily on transcriptomic data [@adamson2014drug]. Epigenomic biomarkers will also become more relevant as several cancer types are increasingly understood to be driven by epigenetic misregulation early in their development [@baylin2006epigenetic]. For example, methylation of the MGMT promoter is a well known biomarker in brain tumors for sensitivity to the standard treatment, temozolomide [@hegi2005mgmt].

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance, only 5 publications in PubMed mentioned BRAF V600E in the title or abstract in 2004 compared to 454 papers in 2017. In order to maintain a high quality and up-to-date knowledge-base, a curation pipeline must be established. This typically involves a queue for papers, triaging those that should be curated and then assignment to a highly experienced curator. This prioritisation step is important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.

Text mining methods have become a common approach to help prioritise literature curation. These methods fall broadly into two main categories, information retrieval (IR) and information extraction (IE). IR methods focus on paper-level information and can take multiple forms. Complex search queries for specific terms or paper metadata (helped by the MeSH term annotations of papers in biomedicine) are common tools for curators. More advanced document clustering and topic modelling systems can use semi-supervised methods to predict whether a paper would be relevant for curation. Examples of this approach include the document clustering method used for the ORegAnno project [@aerts2008text].

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the miRTex knowledgebase, which collates microRNAs and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases. Our previous work has used information extraction methods to extract the role of genes in cancer but did not identify specific aberrations or the clinical relevance of them [@lever2019cancermine].

@@

The full PubMed, PubMed Central Open Access (PMCOA) subset and PubMed Author Manuscript Collection (PMCAMC) corpora were downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as "Introduction", "Methods", "Results" and many synonyms of these  (accessible through the GitHub repository). The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. The updated files from PubMed were processed to identify the latest version of each abstract to process.

@@

Term lists were curated for genes, diseases and drugs based on several resources. The cancer list was curated from a section of the Disease Ontology [@schriml2011disease]. All terms under the "cancer" (DOID:162) parent term were selected and filtered for nonspecific names of cancer (e.g. "neoplasm" or "carcinoma")  . These cancer types were then matched with synonyms from the Unified Medical Language System (UMLS) Metathesaurus [@bodenreider2004unified] (2017AB), either through existing external reference links in the Disease Ontology or through exact string-matching on the main entity names. The additional synonyms in the UMLS were then added through this link. The genes list was built from the Entrez gene list and complemented with UMLS terms. Terms that overlapped with common words found in scientific literature (e.g. ice) were removed.

The drug list was curated from the WikiData resource [@vrandevcic2014wikidata]. All Wikidata entities that are drug instances (Wikidata identifier: Q12140) were selected using a SPARQL query. The generic name, brand name and synonyms were extracted where possible. This list was complemented by a custom list of general drug categories (e.g. chemotherapy, tyrosine kinase inhibitors, etc) and a list of inhibitors built using the previously discussed gene list. This allowed for the extraction of terms such as "EGFR inhibitors". This was done because analysts are often interested in and publications often discuss biomarkers associated with drug classes that target a specific gene, in addition to specific drugs.

All term lists were filtered with a stopwords list. This was based on the stopword list from the Natural Language Toolkit [@bird2006nltk] and the most frequent 5,000 words found in the Corpus of Contempory American English [@davies2009385] as well as a custom set of terms  . It was then merged with common words that occur as gene names (such as ICE).

@@

The word lists and tools used to generate them are accessible through the BioWordlists project (https://github.com/jakelever/biowordlists) and data can be found in the Zenodo repository ((http://doi.org/10.5281/zenodo.1286661).

@@

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. For this work, it was changed to Spacy [@honnibaljohnson] for the improved Python bindings in version 2 for this project.   This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files.

Exact string matching was then used against the tokenized sentences to extract mentions of cancer types, genes, drugs and variants. Longer terms were prioritised during extraction so that "non small cell lung cancer" would be extracted instead of just "lung cancer". Variants were also extracted with a regular expression system for extracting protein coding point mutations (e.g. V600E).

Gene fusions (such as *BCR-ABL1*) were detected by identifying mentions of genes separated by a forward slash, hyphen or colon. If the two entities had no overlapping HUGO IDs, then it was flagged as a possible gene fusion and combined into a single entity. If there were overlapping IDs, it was deemed likely to be referring to the same gene. An example is *HER2/neu* which is frequently seen and refers to a single gene (*ERBB2*) and not a gene fusion.

Acronyms were also detected, where possible, by identifying terms in parentheses and checking the term before it, for instance "non-small cell lung carcinoma (NSCLC)". This was done to remove entity mistakes where possible. The acronym detection method takes the short form (the term in brackets) and iterates backwards through the long form (the term before brackets) looking for potential matches for each letter. If the long form and short form has overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If only one of the long form or short form has an associated ontology ID, they are combined and assigned the associated ontology ID. If both long form and short form have ontology IDs but there is no overlap, the short form is disregarded as the long form has more likelihood of getting the specific term correct  .

@@

knitr::kable(
    keywords, 'pandoc',
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms used to identify sentences that potentially discussed the four evidence types  . Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
    )
```

With all biomedical documents parsed and entities tagged, all sentences were selected that mention at least one gene, at least one cancer and at least one variant. A drug was not required as only one (predictive) of the four evidence types involves a drug entity. These sentences were enriched by filtering with certain keywords that are strongly associated with the different evidence items. The full list and groupings of keywords are shown in Table \@ref(tab:keywords)). This grouping is done to make sure that each evidence type is represented reasonably equally in the training data. The General category with the keyword "marker" is included to catch additional sentences that discuss markers, which may relate to any of the four evidence types. Several of the keywords are stems in order to capture different forms of the word, e.g. prognosis or prognostic. The acronym "DFS" which means "disease free survival"  is also included as it was found in many sentences describing prognosis.

@@

A web platform for simple relation annotation was built using Bootstrap (https://getbootstrap.com/). This allowed annotators to work using a variety of devices, including their smartphones. The annotation system could be loaded with a set of sentences with entity annotations stored in a separate file (also known as standoff annotations). When provided with a relation pattern, for example "Gene/Cancer", the system would search the input sentences and find all pairs of the given entity types in the same sentence. It would make sure that the two entities are not the same term, as in some sentences a token (or set of tokens) could be annotated as both a gene name and a cancer type (e.g. "retinoblastoma"). For a sentence with 2 genes and 2 cancer types, it would find all four possible pairs of gene and cancer type.

```{r annotationplatform, echo=F, eval=T, fig.width=8, fig.asp=0.4,  fig.cap='(ref:annotationplatform)'}
knitr::include_graphics("civicmine/fig_annotationScreenshot.png")
```

(ref:annotationplatform) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown describes a prognostic marker.

Each sentence, with all the possible candidate relations matching the relation pattern, would be presented to the user, one at a time (Fig \@ref(fig:annotationplatform)). The user can then select various toggle buttons for the type of relation that these entities are part of. They can also use these to flag entity extraction errors or mark contentious sentences for discussion with other annotators.

@@

```{r annotationOverview, echo=F, eval=T, fig.width=8, fig.asp=0.4,  fig.cap='(ref:annotationOverview)'}
knitr::include_graphics("civicmine/fig_annotationOverview.png")
```

(ref:annotationOverview) An overview of the annotation process. Sentences are identified from the literature that describe cancers, genes, variants and optionally drugs before being filtered using search terms. The first test phase tried complex annotation of biomarker and variants together but was unsuccessful. The annotation task was split into two separate tasks for biomarkers and variants separately. Each task had a test phase and then the main phase on the 800 sentences that were used to create the gold set.

For the annotation step (outlined in Fig \@ref(fig:annotationOverview)), the annotated data set (known as the gold set) was constructed using a consensus of multiple annotators. An equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types as otherwise the prognostic type dominated the other groups. If this step was not done, 100 randomly selected sentences would only contain 2 (on average) from the diagnostic group. However, this sampling provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. All together, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

Sentences that contain many permutations of relationships (e.g. a sentence with 6 genes and 4  cancer types mentioned) were removed. An upper limit of 5 possible relations was enforced for each sentence. This was done with the knowledge that the subsequent relation extraction step would have a greater false positive rate for sentences with very large number of possible relations. It was also done to make the annotation task more manageable. An annotation manual was constructed with examples of sentences that would and would not match the four evidence types. This was built in collaboration with CIViC curators and is available in our Github repository (https://github.com/jakelever/civicmine).

Each annotation task began with a test phase of 100 sentences. This allows the annotators to become accustomed to the annotation platform and make adjustments to the annotation manual to clarify misunderstandings. The first test phase (Biomarker + Variant) involved annotating sentences for ternary (gene, cancer, variant) or quaternary (gene, cancer, variant, drug) relationships. The ternary relationships included diagnostic, prognostic and predisposing and the quaternary relationship was predictive. A low F1-score inter-annotator agreement (average of 0.52) forced us to reconsider the annotation approach. From our evaluation, we determined that this poor agreement was likely due to including variants within the annotations. This created a large combinatorial problem where it was often unclear exactly which entity mentions to include within a relationship. In order to simplify the problem, the task was split into two separate annotation tasks, the biomarker annotation and the variant annotation. The biomarker annotation involved binary (gene, cancer) and ternary (gene, cancer, drug) relations that described one of the evidence types. The predictive and prognostic evidence types were merged (as shown in Figure 2), to further reduce the annotation complexity. The predictive/prognostic annotations could be separated after tagging as relationships containing a drug would be predictive and those without would be prognostic.   A further postprocessing step to generate the gold set involved identifying prognostic relationships that overlapped with predictive relationships (i.e. shared the same gene and cancer type in a sentence) and removing them. The variant annotation task (gene, variant) focused on whether a variant (e.g. deletion) was associated with a specific gene in the sentence.

With the redefined annotation task, six annotators were involved in biomarker annotation, all with knowledge of the CIViC platform and having experience interpreting patient cancer variants in a clinical context. Three annotators (one of whom was involed in the biomarker annotation)   were involved in variant annotationand they all had experience in cancer genomics. Both annotation tasks started with a new 100-sentence test phase to evaluate the redefined annotation tasks and resolve any ambiguity within the annotation manuals. Good inter-annotator agreement was achieved at this stage for both the biomarker annotation (average F1-score = 0.68) and variant annotation (average F1-score = 0.95). These 100 sentences were discarded as they exhibited a learning curve as annotators become comfortable with the task.

>>>>

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_overview.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
source('civicmine/fig_evaluationresults.R')
source('civicmine/fig_germlineOrSomatic.R')
source('civicmine/fig_furthercomparisons.R')
invisible(dev.off())

@@

**Results**: We extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full-text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs, and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications.

@@

The ability to stratify patients into groups that are clinically related is an important step towards a personalized approach to cancer. Over time, a growing number of biomarkers have been developed to select patients who are more likely to respond to certain treatments. These biomarkers have also been valuable for prognostic purposes and for understanding the underlying biology of the disease by defining different molecular subtypes of cancers that should be treated in different ways (e.g. *ERBB2*/*ESR1*/*PGR* testing in breast cancer [@onitilo2009breast]). Immunohistochemistry techniques are a primary approach for testing samples for diagnostic markers (e.g. CD15 and CD30 for Hodgkin’s disease [@rudiger1998differential]). Recently, the lower cost and increased speed of genome sequencing have also allowed the DNA and RNA of individual patient samples to be characterized for clinical applications [@prasad2016precision]. Throughout the world, this technology is beginning to inform clinician decisions on which treatments to use [@shrager2014rapid]. Such efforts are dependent on a comprehensive and current understanding of the clinical relevance of variants. For example, the Personalized Oncogenomics project at BC Cancer identifies somatic events in the genome such as point mutations, copy number variations, and large structural changes and, in conjunction with gene expression data, generates a clinical report to provide an ‘omic picture of a patient’s tumor [@laskin2015lessons].

@@

Several groups have built in-house knowledgebases, which are developed as analysts examine increasing numbers of cancer patient samples. This tedious and largely redundant effort represents a substantial interpretation bottleneck impeding the progress of precision medicine [@good2014organizing]. To encourage a collaborative effort, the CIViC knowledgebase (https://civicdb.org) was launched to provide a wiki-like, editable online resource where community-contributed edits and additions are moderated by experts to maintain high-quality variant curation [@griffith2017civic]. The resource provides information about clinically-relevant variants in cancer described in the peer-reviewed literature. Variants include protein-coding point mutations, copy number variations, epigenetic marks, gene fusions, aberrant expression levels, and other ‘omic events. It supports four types of evidence associating biomarkers with different classes of clinical relevance (also known as evidence types).

@@

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an information extraction-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full-text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full-text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

Several groups have created knowledgebases to aid clinical interpretation of cancer genomes, many of whom have joined the Variant Interpretation for Cancer Consortium (VICC, http://cancervariants.org/). VICC is an initiative that aims to coordinate variant interpretation efforts and, to this end, has created a federated search mechanism to allow easier analysis across multiple knowledgebases [@wagner2018harmonized]. The CIViC project is co-leading this effort along with OncoKB [@chakravarty2017oncokb], the Cancer Genome Interpreter [@tamborero2018cancer], Precision Medicine Knowledge base [@huang2017cancer], Molecular Match, JAX-Clinical Knowledge base [@patterson2016clinical] and others.

Most of these projects focus on clinically-relevant genomic events, particularly point mutations, and provide associated clinical information tiered by different levels of evidence. Only CIViC includes RNA expression-based biomarkers. These may be of particular value for childhood cancers which are known to be ‘genomically quiet’, having accrued very few somatic mutations. Consequently, their clinical interpretation may rely more heavily on transcriptomic data [@adamson2014drug]. Epigenomic biomarkers will also become more relevant as several cancer types are increasingly understood to be driven by epigenetic misregulation early in their development [@baylin2006epigenetic]. For example, methylation of the MGMT promoter is a well-known biomarker in brain tumors for sensitivity to the standard treatment, temozolomide [@hegi2005mgmt].

The literature on clinically relevant cancer mutations is growing at an extraordinary rate. For instance, only 5 publications in PubMed mentioned BRAF V600E in the title or abstract in 2004 compared to 454 papers in 2017. In order to maintain a high-quality and up-to-date knowledge-base, a curation pipeline must be established. This typically involves a queue for papers, a triage system and then assignment to a highly experienced curator. This prioritisation step is important given the limited time of curators and the potentially vast number of papers to be reviewed. Prioritisation must identify papers that contain knowledge that is of current relevance to users of the knowledgebase. For instance, selecting papers for drugs that are no longer clinically approved would not be valuable to the knowledgebase.

Text mining methods have become a common approach to help prioritise literature curation. These methods fall broadly into two main categories, information retrieval (IR) and information extraction (IE). IR methods focus on paper-level information and can take multiple forms. Complex search queries for specific terms or paper metadata (helped by the MeSH term annotations of papers in biomedicine) are common tools for curators. More advanced document clustering and topic modelling systems can use semi-supervised methods to predict whether a paper would be relevant to curation. Examples of this approach include the document clustering method used for the ORegAnno project [@aerts2008text].

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs, and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant to the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the miRTex knowledgebase, which collates microRNAs and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases. Our previous work has used information extraction methods to extract the role of genes in cancer but did not identify specific aberrations or the clinical relevance of them [@lever2019cancermine].

@@

The full PubMed, PubMed Central Open Access (PMCOA) subset and PubMed Author Manuscript Collection (PMCAMC) corpora were downloaded from the NCBI FTP website using the PubRunner infrastructure [@anekalla2017pubrunner]. These documents were converted to the BioC format for processing with the Kindred package [@lever2017painless]. HTML tags were stripped out and HTML special characters converted to Unicode. Metadata about the papers were retained including PubMed IDs, titles, journal information, and publication date. Subsections of the paper were extracted using a customised set of acceptable section headers such as "Introduction", "Methods", "Results" and many synonyms of these (accessible through the GitHub repository). The corpora were downloaded in bulk in order to not overload the EUtils RESTFUL service that is offered by the NCBI. The updated files from PubMed were processed to identify the latest version of each abstract to process.

@@

Term lists were curated for genes, diseases and drugs based on several resources. The cancer list was curated from a section of the Disease Ontology [@schriml2011disease]. All terms under the "cancer" (DOID:162) parent term were selected and filtered for nonspecific names of cancer (e.g. "neoplasm" or "carcinoma"). These cancer types were then matched with synonyms from the Unified Medical Language System (UMLS) Metathesaurus [@bodenreider2004unified] (2019AA), either through existing external reference links in the Disease Ontology or through exact string-matching on the main entity names. The additional synonyms in the UMLS were then added through this link. The genes list was built from the Entrez gene list and complemented with UMLS terms. Terms that overlapped with common words found in scientific literature (e.g. ice) were removed.

The drug list was curated from the WikiData resource [@vrandevcic2014wikidata]. All Wikidata entities that are medication instances (Wikidata identifier: Q12140) were selected using a SPARQL query. The generic name, brand name, and synonyms were extracted where possible. This list was complemented by a custom list of general drug categories (e.g. chemotherapy, tyrosine kinase inhibitors, etc) and a list of inhibitors built using the previously discussed gene list. This allowed for the extraction of terms such as "EGFR inhibitors". This was done because analysts are often interested in and publications often discuss biomarkers associated with drug classes that target a specific gene.

All term lists were filtered with a stopwords list. This was based on the stopword list from the Natural Language Toolkit [@bird2006nltk] and the most frequent 5,000 words found in the Corpus of Contemporary American English [@davies2009385] as well as a custom set of terms. It was then merged with common words that occur as gene names (such as ICE).

@@

The word lists and tools used to generate them are accessible through the BioWordlists project (https://github.com/jakelever/biowordlists) and data can be found in the Zenodo repository (http://doi.org/10.5281/zenodo.1286661).

@@

The BioC corpora files were processed by the Kindred package. This NLP package used Stanford CoreNLP [@manning2014stanford] for processing in the original published version [@lever2017painless]. For this work, it was changed to Spacy [@honnibaljohnson] for the improved Python bindings in version 2 for this project. This provided easier integration and execution on a cluster without running a Java subprocess. Spacy was used for sentence splitting, tokenization and dependency parsing of the corpora files. Furthermore, we use the Scispacy parsing model [@Neumann2019ScispaCyFA].

Exact string matching was then used against the tokenized sentences to extract mentions of cancer types, genes, drugs, and variants. Longer terms were prioritised during extraction so that "non small cell lung cancer" would be extracted instead of just "lung cancer". Variants were also extracted with a regular expression system for extracting protein-coding point mutations (e.g. V600E).

Gene fusions (such as *BCR-ABL1*) were detected by identifying mentions of genes separated by a forward slash, hyphen or colon. If the two entities had no overlapping HUGO IDs, then it was flagged as a possible gene fusion and combined into a single entity. If there were overlapping IDs, it was deemed likely to be referring to the same gene. An example is *HER2/neu* which is frequently seen and refers to a single gene (*ERBB2*) and not a gene fusion. We used the 24 gene fusions associated with acute myeloid leukemia from MyCancerGenome (https://www.mycancergenome.org/) as a sanity check and found that 23 were found in the literature using this method with only RPN1-MECOM missing.

Acronyms were also detected, where possible, by identifying terms in parentheses and checking the term before it, for instance, "non-small cell lung carcinoma (NSCLC)". This was done to remove entity mistakes where possible. The acronym detection method takes the short-form (the term in brackets) and iterates backward through the long-form (the term before brackets) looking for potential matches for each letter. If the long-form and short-form have overlapping associated ontology IDs, they likely refer to the same thing and can be combined, as in the example above. If only one of the long-form or short-form has an associated ontology ID, they are combined and assigned the associated ontology ID. If both long-form and short-form have ontology IDs but there is no overlap, the short-form is disregarded as the long-form has more likelihood of getting the specific term correct.

@@

knitr::kable(
    keywords, 'pandoc',
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms used to identify sentences that potentially discussed the four evidence types. Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
    )
```

With all biomedical documents parsed and entities tagged, all sentences were selected that mention at least one gene, at least one cancer and at least one variant. A drug was not required as only one (predictive) of the four evidence types involves a drug entity. We evaluated 100 randomly selected sentences and found that only 10 contained information potentially relevant to CIViC, with 7 of the sentences referring to prognostic associations. Many of the sentences report genetic events found in cancer types, methods and other irrelevant information. Manual annotation of a dataset with only 10% relevance would be hugely inefficient and frustrating for expert annotators. Furthermore, any machine learning system would face a large challenge dealing directly with a class balance of 10%. Therefore we elected to use a keyword search to enrich the sentences with CIViC relevant knowledge.

Through manual review of a subset of the sentence combined with knowledge of the requirement of CIViC, we selected the keywords found in Table \@ref(tab:keywords). Most of the keywords target a specific association type (e.g. survival for prognostic). This set was not designed to be exhaustive but to keep a reasonable balance of relevant sentences that could be later filtered by a machine learning system. In selecting each keyword, the filtered sentences were evaluated for relevance and the keyword was added if at least half of the sentences seemed relevant to CIViC. The five groups were treated separately such that 20% of the corpus comes from each of the five groups. This was done to provide coverage for the rarer types such as diagnostic that were not found at all in the initial 100 sentences evaluated.

@@

```{r annotationToolAndOverview, echo=F, eval=T, fig.width=18, fig.asp=0.4, fig.cap='(ref:annotationToolAndOverview)'}

annotationScreenshot <- rasterGrob(readPNG(source="civicmine/fig_annotationScreenshot.png"))

annotationOverview <- rasterGrob(readPNG(source="civicmine/fig_annotationOverview.png"))

annotationScreenshot.withTitle <- arrangeGrob(annotationScreenshot,top='(a)')
annotationOverview.withTitle <- arrangeGrob(annotationOverview,top='(b)')

grid.arrange(annotationScreenshot.withTitle,
             annotationOverview.withTitle, 
             ncol=2)
```

(ref:annotationToolAndOverview) (a) A screenshot of the annotation platform that allowed expert annotators to select the relation types for different candidate relations in all of the sentences. The example sentence shown describes a prognostic marker. (b) An overview of the annotation process. Sentences are identified from literature that describes cancers, genes, variants and optionally drugs before being filtered using search terms. The first test phase tried complex annotation of biomarker and variants together but was unsuccessful. The annotation task was split into two separate tasks for biomarkers and variants separately. Each task had a test phase and then the main phase on the 800 sentences that were used to create the gold set.

A web platform for simple relation annotation was built using Bootstrap (https://getbootstrap.com/). This allowed annotators to work using a variety of devices, including their smartphones. The annotation system could be loaded with a set of sentences with entity annotations stored in a separate file (also known as standoff annotations). When provided with a relation pattern, for example, "Gene/Cancer", the system would search the input sentences and find all pairs of the given entity types in the same sentence. It would make sure that the two entities are not the same term, as in some sentences a token (or set of tokens) could be annotated as both a gene name and a cancer type (e.g. "retinoblastoma"). For a sentence with 2 genes and 2 cancer types, it would find all four possible pairs of gene and cancer type.

Each sentence, with all the possible candidate relations matching the relation pattern, would be presented to the user, one at a time (Fig \@ref(fig:annotationToolAndOverview)a). The user can then select various toggle buttons for the type of relation that these entities are part of. They can also use these to flag entity extraction errors or mark contentious sentences for discussion with other annotators.

@@

For the annotation step (outlined in Fig \@ref(fig:annotationToolAndOverview)b), the annotated data set (known as the gold set) was constructed using a consensus of multiple annotators. An equal number of sentences were selected from each of the groups outlined in Table \@ref(tab:keywords). This guaranteed coverage of all four evidence types as otherwise the prognostic type dominated the other groups. If this step was not done, 100 randomly selected filtered sentences would only contain 2 (on average) from the diagnostic group. However, this sampling provided poor coverage of sentences that describe specific point mutations. Many precision oncology projects only focus on point mutations and so a further requirement was that 50% of sentences for annotation include a specific point mutation. Altogether, this sampling provides better coverage of the different omic events and evidence types that were of interest. Special care is required when evaluating models built on this customized training set as an unweighted evaluation would not be representative of the real literature.

Sentences that contain many permutations of relationships (e.g. a sentence with 6 genes and 4 cancer types mentioned) were removed. An upper limit of 5 possible relations was enforced for each sentence. This was done with the knowledge that the subsequent relation extraction step would have a greater false positive rate for sentences with a very large number of possible relations. It was also done to make the annotation task more manageable. An annotation manual was constructed with examples of sentences that would and would not match the four evidence types. This was built in collaboration with CIViC curators and is available in our Github repository (https://github.com/jakelever/civicmine). Each annotation task began with a test phase of 100 sentences. This allows the annotators to become accustomed to the annotation platform and make adjustments to the annotation manual to clarify misunderstandings.

The first test phase (Biomarker + Variant) involved annotating sentences for ternary (gene, cancer, variant) or quaternary (gene, cancer, variant, drug) relationships. The ternary relationships included diagnostic, prognostic and predisposing and the quaternary relationship was predictive. As many sentences contain multiple mentions of the same gene or variant, we found there was a combinatorial problem as different annotators found it challenging to decide which variants should be associated with which gene. The annotators were trying to decide linguistically which of the mentions was part of the biomarker being described. For example in a sentence that mentioned the same variant five times, different annotators chose different mentions of the same variant. These were flagged as differences and reduced the annotator agreement. This led to the low F1-score inter-annotator agreement (average of 0.52) and forced us to reconsider the annotation approach.

To reduce the possible combinations, we split the task into two separate tasks, the biomarker annotation and the variant annotation. The biomarker annotation involved binary (gene, cancer) and ternary (gene, cancer, drug) relations that described one of the evidence types. The variant annotation task (gene, variant) focused on whether a variant (e.g. deletion) was associated with a specific gene in the sentence. For a sentence containing two genes, two cancer types, and three variants, the original combined task would have twelve combinations that would require annotation. By splitting it into the two tasks, the biomarker task would have four combinations and the variant task would also have four combinations. We hypothesised that a smaller number of combinations would reduce the cognitive load for the annotators and increase inter-annotator agreement. To further reduce complexity, the predictive and prognostic evidence types were merged (as shown in Figure 2), to further reduce the annotation complexity. The predictive/prognostic annotations could be separated after tagging as relationships containing a drug would be predictive and those without would be prognostic. A further postprocessing step to generate the gold set involved identifying prognostic relationships that overlapped with predictive relationships (i.e. shared the same gene and cancer type in a sentence) and removing them.

With the redefined annotation task, six annotators were involved in biomarker annotation, all with knowledge of the CIViC platform and having experience interpreting patient cancer variants in a clinical context. Three annotators (one of whom was involved in the biomarker annotation) were involved in variant annotation and they all had experience in cancer genomics. Both annotation tasks started with a new 100-sentence test phase to evaluate the redefined annotation tasks and resolve any ambiguity within the annotation manuals. Good inter-annotator agreement was achieved at this stage for both the biomarker annotation (average F1-score = 0.68) and variant annotation (average F1-score = 0.95). The higher agreement scores validated our reasoning to split the annotation task in two. In fact, the very high variant annotation score suggests that this task was made relatively easy by separating it. These 100 sentences were discarded as they exhibited a learning curve as annotators become comfortable with the task. Between each annotation stage, the annotators discussed through video conference the difficulties that had been encountered. These comments were used to improve the annotation manuals with the aim to capture sentences with greater relevance to CIViC and also increase inter-annotator agreement.
